---
title: "Severe Weather Events in the U.S. From 1950 to 2011"
author: "Jose A. Mora Soto"
output: html_document
---

# Synopsis
Severe weather events are one of the mayor disasters in the U.S. provoking several
mortal victims and mayor economic losses. This report summarizes the information
from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database
showing information about the most fatal weather events in terms of mortality
and economic consequences.

# Data Processing

## Data Source
The data used for this report is publicly offered by NOAA for any analysis and
research purposes; the specific data for this report contains information about
the severe weather events that occured in the U.S. from 1950 to November 2011,
this data file was provided as a resource in the **Reproducibe Research** course
from the Jonh Hopkins University at Coursera for the second course project, it
can be downloaded from this [web link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

## Data Preparation
For practicity the data file was downloaded and uncompresed using the operating
system's tools and renamed as _StormData.csv_.

The first step to conduct the analysis was to load the database into a data frame
as follows:


```{r}

```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
