---
title: "Severe Weather Events in the U.S. From 1950 to 2011"
author: "Jose A. Mora Soto"
output: html_document
---

```{r libraries}
# Libraries needed
library(data.table)
library(pander)
```


# Synopsis
Severe weather events are one of the mayor disasters in the U.S. provoking
several mortal victims and mayor economic losses. This report summarizes the
information from the U.S. National Oceanic and Atmospheric Administration's
(NOAA) storm database showing information about the most fatal weather events in
terms of mortality and economic consequences.

# Data Processing

## Data Source
The data used for this report is publicly offered by NOAA for any analysis and
research purposes; the specific data for this report contains information about
the severe weather events that occured in the U.S. from 1950 to November 2011,
this data file was provided as a resource in the **Reproducibe Research** course
from the Jonh Hopkins University at Coursera for the second course project, it
can be downloaded from this [web link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

## First Steps Processing the Data
For practicity the data file was downloaded and uncompresed using the operating
system's tools and renamed as _StormData.csv_.

The first step to conduct the analysis was to load the database into a data
frame as follows:

```{r data_load, cache = T}
# The data will be retrieved into the "stormdata" data frame, since the data
# file is big, we check if it has already loaded.
if (!exists("stormdata")) {
  stormdata <- read.csv("StormData.csv")
}
```

Once the data was loaded in the __stormdata__ dataframe, some data preparation
was done by creating some data structures to manipulate the data to create the
report. The most relevant data structures defined were the following:

* __stormdata_dt__: A data table structure (it needs the __data.table__ package)
  to store all the data from the original file and have extra funcitions to
  clean and prepare the data.
* __harmful_events__: A data table structure to store the data needed to answer
  Question 1 of the project project.
* __economic_impact__: A data frame structure to store the data needed to answer
  Question 2 of the course project.

## Data Processing for Question 1
In order to answer Question 1 __Across the United States, which types of events
(as indicated in the **EVTYPE** variable) are most harmful with respect to
population health?__ it has been considered that the most harmful situation is
dead, so the number of fatalities per severe weather event were used as
indicator. The data was retrieved and stored in the __harmful_events__ data
table as follows:

```{r data_preparation_question_1}
# Data preparatrion
stormdata_dt <- data.table(stormdata)
stormdata_dt$EVTYPE <- toupper(stormdata_dt$EVTYPE)
# Sum all FATALITIES per EVTYPE
harmful_events <- stormdata_dt[, sum(FATALITIES), by = EVTYPE]
harmful_events <- harmful_events[order(-V1),]
names(harmful_events) = c("EventType", "TotalFatalities")
harmful_events <- harmful_events[
  which(harmful_events$TotalFatalities > 0 & 
          substr(harmful_events$EventType, 1, 7) != "SUMMARY")
  ]
```

## Data Processing for Question 2
In order to ansquer Question 2 __Across the United States, which types of events
have the greatest economic consequences?__ the data was retrieved and stored in
the __economic_impact__ data frame; later the economic loss was calculated by
losses in crops as well as in properties as it is shown in the code bellow.

```{r data_preparation_question_2, cache = T}
# Data preparation
exponent <- c('H' = 100, 'K' = 1000, 'M' = 1000000, 'B' = 1000000000)
if (!exists("economic_impact")) {
 # Creating the economic_impact dataframe
  economic_impact <- data.frame(
                                "EVTYPE" = stormdata_dt$EVTYPE,
                                "PROMDMG" = stormdata_dt$PROPDMG,
                                "PROPDMGEXP" = toupper(stormdata_dt$PROPDMGEXP),
                                "CROPDMG" = stormdata_dt$CROPDMG,
                                "CROPDMGEXP" = toupper(stormdata_dt$CROPDMGEXP)
                                )
  # Calculating the monetary impact of each damage
  economic_impact$PROPDMGCASH <- economic_impact$PROMDMG * 
                              exponent[as.character(economic_impact$PROPDMGEXP)]
  economic_impact$CROPDMGCASH <- economic_impact$CROPDMG *
                              exponent[as.character(economic_impact$CROPDMGEXP)]

# Top 10 most costly events by properties and crops
# Economic Loss in properties
economic_impact_prop <- aggregate(
  economic_impact$PROPDMGCASH ~ economic_impact$EVTYPE, 
  economic_impact, 
  sum
)
names(economic_impact_prop) <- c("EVTYPE", "PROPDMGCASH")
economic_impact_prop <- 
  economic_impact_prop[order(-economic_impact_prop$PROPDMGCASH),]

# Economic Loss in crops
economic_impact_crop <- aggregate(
  economic_impact$CROPDMGCASH ~ economic_impact$EVTYPE,
  economic_impact,
  sum
)
names(economic_impact_crop) <- c("EVTYPE", "CROPDMGCASH")
economic_impact_crop <- 
  economic_impact_crop[order(-economic_impact_crop$CROPDMGCASH),]
}
```

# Results
Once the data was processed it is possible to present some results about the
impact of the severe weather events in the U.S.

## Question 1 __Across the United States, which types of events
(as indicated in the **EVTYPE** variable) are most harmful with respect to
population health?__

Bellow in Table 1 a list of the top 10 most harmful severe weather events in
the U.S. is presented, as can be seen __Tornados__ are the events with the
greater number of fatalities from 1950 to 2011.

```{r table_top_10_fatal_events, cache = T}
panderOptions("digits", 2)
set.caption("Table 1: Most Harmful Events in the U.S.
            (by number of fatalities)")
pander(head(harmful_events, 10))
```

Following a bar plot is presented to graphicaly depict the results from Table 1.

```{r barplot_top_10_fatal_events}
par(las = 2) # Labels are perpendicular to the axis
barplot(head(harmful_events$TotalFatalities, 10),
  main = "Top 10 Harmful Events in the U.S. From 1950 to 2011 (by Fatalities)",
  names.arg = harmful_events$EventType[1:10],
  ylab = "Number of Fatalities",
  col = rainbow(10)
  )
set.caption("Figure 1: Most Harmful Events in the U.S. (by number of
            fatalities)")
```

## Question 2:  __Across the United States, which types of events have the
greatest economic consequences?__

To answer this question the data was splited in two dimensions, one to analyze
the economic loss in crops and another to analyze the economic loss in
properties (as presented above in the the __Data Processing__ section). Next in
Figure 2 a pie chart showing the percentage of the total economic losses that
were suffered in crops and propierties is presented.

```{r pie_total_economic_loss}
slices = c(sum(economic_impact$PROPDMGCASH, na.rm = T),
           sum(economic_impact$CROPDMGCASH, na.rm = T))
lbls = c("Properties", "Crops")
pct = round(slices/sum(slices)*100)
lbls = paste(lbls, 
             format(slices,
                    decimal.mark = ".",
                    big.mark = " ",
                    small.mark = ",",
                    small.interval = 3),
             sep = "\n$ ") # add the money
lbls = paste(lbls, "\n(")
lbls = paste(lbls, pct, sep = "")
lbls = paste(lbls, "%)", sep = "")
pie(
  slices,
  lbls,
  main = "Total Economic Loss From All The Severe Weather Events (USD)",
  col = c("Yellow", "Orange")
  )
set.caption("Figure 2: Total Economic Loss From All The Severe Weather Events
            (USD)")
```

Finally in Figure 3 a bar chart is presented in Figure 3 to present the top 10
most costly severe events in terms of losses in properties.

```{r barplot_top_10_most_costly_events_properties}
par(las = 2)
barplot(head(economic_impact_prop$PROPDMGCASH, 10),
        main = "Top 10 Most Costly Severe Weather Events in the U.S. 
        From 1950 to 2011 (Properties)",
        names.arg = economic_impact_prop$EVTYPE[1:10],
        ylab = "USD",
        col = rainbow(10)
)
set.caption("Figure 3: Top 10 Most Costly Severe Weather Events in the U.S.
            From 1950 to 2011 (Properties)")
```

# Full Code to Reproduce this Analysis
If you want to see and play yourself with the code I made to create this report,
you can download it at my [GitHub code repository for this project]
(https://github.com/jarturomora/datasciencecoursera/tree/master/course5/courseproject2).

In case you want to contact me, [feel free to contact me](https://about.me/jarturomora).
