---
title: "Quiz 2 - Rationale"
author: "Jose Arturo Mora Soto"
date: "April 21, 2016"
output: html_document
---

## Remarks
I made this document for my own understanding of the responses thayt I gave to
the questions of this quiz.

## Question 1
What is the variance of the distribution of the average an IID draw of _n_
observations from a population with mean $\mu$.

### Answer 1
From the slides of topic 7 "Variability", on slide 9 it is explained that the
variance of the sample ($S^2$) from a population can be calculated using the
population variance ($\sigma^2$) using the formula: 
$$S^2 = \frac{\sigma^2}{n}$$

## Question 2
Suppose that diastolic blood pressures (DBPs) for men aged 35-44 are normally
distributed with a mean of 80 (mm Hg) and a standard deviation of 10. About what
is the probability that a random 35-44 year old has a DBP less than 70?

### Answer 2
Since the data is normally distributed we can calculate this probability using
the ```pnorm()``` function as follows:

``` {r}
dbp_quantile = 70
dbp_mean = 80
dbp_sd = 10
pnorm(dbp_quantile, dbp_mean, dbp_sd)
```

## Question 3
Brain volume for adult women is normally distributed with a mean of about
1,100 cc for women with a standard deviation of 75 cc. About what brain volume
represents the 95th percentile?

### Answer 3
Using the given information about the scenario, we deduce that we need to find
the quantile that represents the 95th percentile; to do that, we can use the
```qnorm()``` function as follows:

```{r}
brain_vol_percentile = 0.95
brain_vol_mean = 1100
brain_vol_sd = 75
qnorm(brain_vol_percentile, brain_vol_mean, brain_vol_sd)
```

## Question 4
Refer to the previous question. Brain volume for adult women is about 1,100 cc
for women with a standard deviation of 75 cc. Consider the sample mean of 100
random adult women from this population. Around what is the 95th percentile of
the distribution of that sample mean?

### Answer 4
Now we have to estimate the quantile for a sample of 100 women, so we need to
find the standard deviation of the population; we will use the formula given
in **Answer 1** for that

```{r}
sample_size = 100
population_sd = 75
sample_sd = population_sd / sqrt(sample_size)
```

Now we will calculate the quantile using the ```qnorm()``` function.

```{r}
qnorm(brain_vol_percentile, brain_vol_mean, sample_sd)
```

## Question 5
You flip a fair coin 5 times, about what's the probability of getting
4 or 5 heads?

### Answer 5
This is a Bernoulli distribution problem since we have a __X__ with a 
binary outcome, 1 for head, 0 for tail. We have to keep in mind that the
propability of having a head (or tail) is 50%. We will suppose that X is a
[binomial random variable](https://onlinecourses.science.psu.edu/stat414/node/68),
so the probability will be given by the formula:
$$P(X = x) = (_{x}^{n})p^{x}(1-p)^{n-x}$$

The probability of having 4 or 5 heads will be the sum of the probability of
having 4 heads plus 5 heads as follows:

```{r}
p_4heads <- choose(5,4) * 0.5^4 * (1 - 0.5)^(5 - 4)
p_5heads <- choose(5,5) * 0.5^5 * (1 - 0.5)^(5 - 5)
p_4and5 <- p_4heads + p_5heads
round(p_4and5, 2)
```

Since we said __X__ is binomial, we can also calculate the probability using
the ```pbinom()``` function as follows:

```{r}
p <- 0.5 # Probability of having a head
n <- 5 # Number of trials (coin flips)
head_quantile <- 3 # 4 or 5 out of 5
p_4and5 <- pbinom(head_quantile, n, p, lower.tail = FALSE)
round(p_4and5, 2)
```

## Question 6
The respiratory disturbance index (RDI), a measure of sleep disturbance, for a
specific population has a mean of 15 (sleep events per hour) and a standard
deviation of 10. **They are not normally distributed**. Give your best estimate
of the probability that a sample mean RDI of 100 people is between 14 and 16
events per hour?

### Answer 6
According to The Central Limit Theorem (CLT) for a large enough sample size
__n__, the distribution of the sample mean $\bar x$ will approach a normal
distribution, so that since we are talking here about the RDI of all the people
we will suppose that the sample mean will be the population mean and that it's
normal (will tend to be). So we can calculate the probability using the
```pnorm()``` function as follows:

```{r}
n = 100
rdi_mu = 15
rdi_sigma = 10 / sqrt(n) # This is the sample's standard error

# Probability of having 14 events per hour
# (area to the left of 14 in the bell curve)
prob14 <- pnorm(14, rdi_mu, rdi_sigma)

# Probability of having 16 events per hour
# (area to the left of 16 in the bell curve)
prob16 <- pnorm(16, rdi_mu, rdi_sigma)

prob_14and16 <- prob16 - prob14
round(prob_14and16, 2)
```

## Question 7
Consider a standard uniform density. The mean for this density is 0.5 and the
variance is 1/12. You sample 1,000 observations from this distribution and take
the sample mean, what value would you expect it to be near?

### Answer 7
According to the Law of Large Numbers and the Central Limit Theorem the sample
mean would be the population mean: **0.5**. However let's probe it using R.

```{r}
n <- 1000
quant <- 0.5
pop_mu <- 0.5
pop_var <- 1/12
sam_sd <- sqrt(1/12) / sqrt(n)

qnorm(quant, pop_mu, sam_sd)
```

## Question 8
The number of people showing up at a bus stop is assumed to be Poisson with
a mean of 5 people per hour. You watch the bus stop for 3 hours. About what’s
the probability of viewing 10 or fewer people?

### Answer 8

```{r}
t <- 3
mu <- 5 # Since in a Poisson distribution lambda = mean = variance
quant <- 10

round(ppois(quant, lambda = mu * t), 2)
```

